\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-10-07 paper.qmd
Added some references

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = -q, s_2 = -qr_2$, where $q$ is the buffer sound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-param-two_curvy (a) illustrates this. The distribution of bin counts for $s_1$ varying between $-0.1-0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{\frac{n}{r_2}}$ are allowed. @fig-param-two_curvy (b) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

By default, when assessing the choice of $b_1$, the total number of bins is measured by the number of **non-empty** bins. This more accurately reflects the hexagon grid relative the MSE than the full number of bins in the grid. It may also be beneficial to remove low count bins also, in the situation where data is clustered or stringy, where the observed data is sparse. In order to decide if this is necessary, you would examine the distribution of bin counts, or the density which puts the counts on a standard scale. If there is something of a gap at low values, this would suggest a potential value to use as a cutoff. Alternatively, one could choose to remove based on a percentile, the bins with density in the lowest 5% of all bins, for example. @fig-param-two_curvy (c) illustrates the effect on the model representation of removing bins below different percentages. Generally, we would urge caution in removing low count bins.

@@

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (@fig-param-two_curvy). The MSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

Edges define the neighbourhood structure, in order to provide a smooth \gD{} representation of the fitted model. @fig-two_curvy-true-sc shows a wire frame of the true model that was used to generate the S-curve example data. The ideal is that the representation of the fitted model, at least for this example where we know the true model, should look similar to this.

@@

The decision on edge length removal is made based on the distribution of edge lengths. In particular, a gap between values, where there a concentration of small values and then a few larger values, likely suggests a cutoff for edge removal. Because the triangulation is typically done on the hexagon centroids, there are particular discrete edge lengths, based on bin widths. @fig-NLDR-two-curvy (d) illustrates edge length distributions.

@@

```{r}
#| fig-cap: "Distribution of edge lengths and wireframes in \\gD{} and $7\\text{-}D$ with Delaunay triangulation, and two choices of long edge removal: (b) benchmark = 4.5$a_1$ and (c) benchmark = 2$a_1$, where $a_1$ = 0.09."
#| fig-height: 15
#| fig-width: 15
#| out-height: 70%

>>>>

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = -q, s_2 = -qr_2$, where $q$ is the buffer bound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-bins-two-curvy illustrates this. The distribution of bin counts for $s_1$ varying between $-0.1-0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{\frac{n}{r_2}}$ are allowed. @fig-param-two-curvy (a) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

By default, when assessing the choice of $b_1$, the total number of bins is measured by the number of **non-empty** bins. This more accurately reflects the hexagon grid relative the MSE than the full number of bins in the grid. It may also be beneficial to remove low count bins also, in the situation where data is clustered or stringy, where the observed data is sparse. In order to decide if this is necessary, you would examine the distribution of bin counts, or the density which puts the counts on a standard scale. If there is something of a gap at low values, this would suggest a potential value to use as a cutoff. Alternatively, one could choose to remove based on a percentile, the bins with density in the lowest 5% of all bins, for example. @fig-param-two-curvy (c) illustrates the effect on the model representation of removing bins below different percentages. Generally, we would urge caution in removing low count bins.

@@

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (@fig-param-two-curvy (c)). The MSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

Edges define the neighbourhood structure, in order to provide a smooth \gD{} representation of the fitted model. @fig-two-curvy-true-proj shows a wire frame of the true model that was used to generate the two C-shaped clusters example data. The ideal is that the representation of the fitted model, at least for this example where we know the true model, should look similar to this.

@@

The decision on edge length removal is made based on the distribution of edge lengths. In particular, a gap between values, where there a concentration of small values and then a few larger values, likely suggests a cutoff for edge removal. Because the triangulation is typically done on the hexagon centroids, there are particular discrete edge lengths, based on bin widths. @fig-rm-lg illustrates edge length distributions.

@@

```{r}
#| fig-cap: "Distribution of edge lengths and wireframes in \\gD{} and $7\\text{-}D$ with Delaunay triangulation, and two choices of long edge removal: (b) benchmark = 4.5$a_1$ and (c) benchmark = 2$a_1$, where $a_1$ = 0.09."
#| fig-height: 15
#| fig-width: 15
#| out-height: 70%
#| fig-label: fig-rm-lg

### 2024-10-07 paper.tex
Added some references

The values \((s_1, \ s_2)\) define the position of the centroid of the
bottom left hexagon. By default, this is at \(s_1 = -q, s_2 = -qr_2\),
where \(q\) is the buffer sound the data. The choice of these values can
have some effect on the distribution of bin counts.
\textbf{?@fig-param-two\_curvy} (a) illustrates this. The distribution
of bin counts for \(s_1\) varying between \(-0.1-0.0\) is shown.
Generally, a more uniform distribution among these possibilities would
indicate that the bins are reliably capturing the underlying
distribution of observations.

@@

The default number of bins \(b=b_1\times b_2\) is computed based on the
sample size, by setting \(b_1=n^{1/3}\), consistent with the
Diaconis-Freedman rule \citep{freedman1981}. The value of \(b_2\) is
determined analytically by \(b_1, q, r_2\). Values of \(b_1\) between
\(2\) and \(b_1 = \sqrt{\frac{n}{r_2}}\) are allowed.
\textbf{?@fig-param-two\_curvy} (b) shows the effect of different
choices of \(b_1\) on the MSE of the fitted model.

@@

By default, when assessing the choice of \(b_1\), the total number of
bins is measured by the number of \textbf{non-empty} bins. This more
accurately reflects the hexagon grid relative the MSE than the full
number of bins in the grid. It may also be beneficial to remove low
count bins also, in the situation where data is clustered or stringy,
where the observed data is sparse. In order to decide if this is
necessary, you would examine the distribution of bin counts, or the
density which puts the counts on a standard scale. If there is something
of a gap at low values, this would suggest a potential value to use as a
cutoff. Alternatively, one could choose to remove based on a percentile,
the bins with density in the lowest 5\% of all bins, for example.
\textbf{?@fig-param-two\_curvy} (c) illustrates the effect on the model
representation of removing bins below different percentages. Generally,
we would urge caution in removing low count bins.

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (\textbf{?@fig-param-two\_curvy}). The MSE
shows a gradual decrease as the benchmark value goes from \(1\) to
\(0\). Evaluating this rate of increase is important. If the increment
is not considerable, the decision might lean towards retaining
low-density hexagons.

@@

Edges define the neighbourhood structure, in order to provide a smooth
\gD{} representation of the fitted model.
\textbf{?@fig-two\_curvy-true-sc} shows a wire frame of the true model
that was used to generate the S-curve example data. The ideal is that
the representation of the fitted model, at least for this example where
we know the true model, should look similar to this.

@@

The decision on edge length removal is made based on the distribution of
edge lengths. In particular, a gap between values, where there a
concentration of small values and then a few larger values, likely
suggests a cutoff for edge removal. Because the triangulation is
typically done on the hexagon centroids, there are particular discrete
edge lengths, based on bin widths. Figure~\ref{fig-NLDR-two-curvy} (d)
illustrates edge length distributions.

>>>>

The values \((s_1, \ s_2)\) define the position of the centroid of the
bottom left hexagon. By default, this is at \(s_1 = -q, s_2 = -qr_2\),
where \(q\) is the buffer bound the data. The choice of these values can
have some effect on the distribution of bin counts.
Figure~\ref{fig-bins-two-curvy} illustrates this. The distribution of
bin counts for \(s_1\) varying between \(-0.1-0.0\) is shown. Generally,
a more uniform distribution among these possibilities would indicate
that the bins are reliably capturing the underlying distribution of
observations.

@@

The default number of bins \(b=b_1\times b_2\) is computed based on the
sample size, by setting \(b_1=n^{1/3}\), consistent with the
Diaconis-Freedman rule \citep{freedman1981}. The value of \(b_2\) is
determined analytically by \(b_1, q, r_2\). Values of \(b_1\) between
\(2\) and \(b_1 = \sqrt{\frac{n}{r_2}}\) are allowed.
Figure~\ref{fig-param-two-curvy} (a) shows the effect of different
choices of \(b_1\) on the MSE of the fitted model.

@@

By default, when assessing the choice of \(b_1\), the total number of
bins is measured by the number of \textbf{non-empty} bins. This more
accurately reflects the hexagon grid relative the MSE than the full
number of bins in the grid. It may also be beneficial to remove low
count bins also, in the situation where data is clustered or stringy,
where the observed data is sparse. In order to decide if this is
necessary, you would examine the distribution of bin counts, or the
density which puts the counts on a standard scale. If there is something
of a gap at low values, this would suggest a potential value to use as a
cutoff. Alternatively, one could choose to remove based on a percentile,
the bins with density in the lowest 5\% of all bins, for example.
Figure~\ref{fig-param-two-curvy} (c) illustrates the effect on the model
representation of removing bins below different percentages. Generally,
we would urge caution in removing low count bins.

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (Figure~\ref{fig-param-two-curvy} (c)). The
MSE shows a gradual decrease as the benchmark value goes from \(1\) to
\(0\). Evaluating this rate of increase is important. If the increment
is not considerable, the decision might lean towards retaining
low-density hexagons.

@@

Edges define the neighbourhood structure, in order to provide a smooth
\gD{} representation of the fitted model.
Figure~\ref{fig-two-curvy-true-proj} shows a wire frame of the true
model that was used to generate the two C-shaped clusters example data.
The ideal is that the representation of the fitted model, at least for
this example where we know the true model, should look similar to this.

@@

The decision on edge length removal is made based on the distribution of
edge lengths. In particular, a gap between values, where there a
concentration of small values and then a few larger values, likely
suggests a cutoff for edge removal. Because the triangulation is
typically done on the hexagon centroids, there are particular discrete
edge lengths, based on bin widths. \textbf{?@fig-rm-lg} illustrates edge
length distributions.
